---
title: "HW Week 11"
output:
  html_notebook: default
  html_document: default
---

## Week 11

### Step 1: Load required libraries


```{r, message=FALSE}
library("data.table")
library("lme4")
library("dplyr")
library("ggplot2")
library("lattice")
```

### Step 2: Load data

```{r}
NLSY_wide <- fread("NLSY.txt", na.strings = ".")
NLSY_wide

```


### Step 3: Create New Column Variables

Create 8 new column variables and remove the original two **m_age** and **age**

```{r}
NLSY_wide <- NLSY_wide %>% 
  mutate(m_age1 = m_age,
         m_age2 = m_age + 2,
         m_age3 = m_age + 4,
         m_age4 = m_age + 6,
         c_age1 = age,
         c_age2 = age + 2,
         c_age3 = age + 4,
         c_age4 = age + 6) %>% 
  select(-c(m_age, age))

NLSY_wide
```

### Step 3: Trasform data format

We now transform the wide format into long format.The folowing function gathers Columns 2 to 5 into rows. In place of the four variable read1-read4, we get a **read_score** variable, the key is *id* and the corresponding value is **read_score**. Similiarly, Columns 10 to 13 collapse into **mother_age** and Columns 14 to 17 collapse into **child_age**. After reshaping the dataset, we remove rows containing missing values.
```{r}
NLSY_long <- NLSY_wide %>% 
  reshape(direction = "long", 
          idvar = c("id"),
          varying = list(c(2:5), c(10:13), c(14:17)), 
          v.names = c("read_score", "mother_age", "child_age"), 
          times = c("1", "2", "3", "4")) %>%
  na.omit()
NLSY_long
```

Shift ages and change variable types

```{r}
NLSY_long <- within(NLSY_long, {
  id <- factor(id)
  boy <- factor(boy)
  time <- factor(time)
  mother_age <- mother_age - 6
  child_age <- child_age - 6
})

```

### Step 4: Plot data

We now plot the data for the first 50 children
```{r}

temp_df <- NLSY_long %>% filter(id %in% NLSY_long$id[1:50])
ggplot(data = temp_df , 
       aes(x = child_age , 
           y = read_score, 
           group = boy)) +
  facet_wrap( ~ id, ncol = 10) + 
  geom_line(aes(colour = boy), size = 0.2) +
  xlab("Age") +
  ylab("Reading Score") +
  ggtitle("NLSY Data") +
  theme(legend.position = "none")
```


### Step 5: Fit Model

Here we fit a mixed effect model with random slopes but fixed intercept.

```{r}
NLSY_fit1 <- lmer( read_score ~ child_age + I(child_age^2) + (0 + child_age + I(child_age^2)|id),
                  data = NLSY_long,
                  REML = FALSE)

summary(NLSY_fit1)
```

Here we fit another model with the uncorrelated slopes. Note the double-bar notation.

```{r}
NLSY_fit2 <- lmer( read_score ~ child_age + I(child_age^2) + (0 + child_age + I(child_age^2)||id),
                  data = NLSY_long,
                  REML = FALSE)

summary(NLSY_fit2)
```

We can compare the two models using the anova function.

```{R}
anova(NLSY_fit2, NLSY_fit1)

```


To extract the variances, simply run the following:

```{R}
(vc <- VarCorr(NLSY_fit1))
print(vc, comp = c("Variance"))

```

### Step 6: Check Model

Plot residuals vs fitted.

```{R}
plot(NLSY_fit1)
```

Plot the random effects. They should have a normal distribution.

```{r}
ranef_betas <- ranef(NLSY_fit1, condVar = TRUE)
dotplot(ranef_betas)
```

To check normality assumption, we use the quantile-quantile plot of a sample against a theoretical distribution.

```{r}
qqmath(ranef_betas)

```

